Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-27 VegOzone.tex
Rework introduction

A substantial body of evidence exists that suggests flux-based metrics, that relate stomatal ozone uptake to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics. This is because they can account for particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. Most of these previous studies have focused on the Mediterranean where soil moisture deficits are thought to be most influential in decoupling concentration from uptake. Relatively few studies have explored the situation in Northern European climates where low temperatures may be considered more likely to influence ozone uptake \citep[e.g.]{iF:Juran2018}.

To estimate the flux-based metric, also referred to as the Phyto-toxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance as described further in Section~\ref{sec:do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} over the growing season.

@@

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. A similar bio-monitoring study was conducted at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd (Fig.~\ref{fig:data_svanvik_2018_2019} but also over whole northern Fennoscandia in July. Therefore conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia in our study.

During the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.\\

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). To this end, we derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.

>>>>

A substantial body of evidence exists that suggests flux-based metrics, that relate stomatal ozone uptake to vegetation damage, are biologically more relevant for risk assessments than exposure-based metrics. This is because they can account for particular species characteristics (i.e. physiology and phenology) as well as environmental conditions that can decouple the relationship between ozone concentration, ozone uptake, and consequent damage \citep{PT:Emberson2020}. These flux-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the climatically extreme parts of Europe \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017}. Most of these previous studies have focused on the Mediterranean where soil moisture deficits are thought to be most influential in decoupling concentration from uptake. Relatively few studies have explored the situation in Northern European climates \citep[e.g.][]{iF:Juran2018}.

To estimate the flux-based metric, also referred to as the Phyto-toxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} over the growing season.

@@

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. We conducted a similar bio-monitoring study at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July. It is very likely that elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions like during the 2003 drought period,\citep{JGR:Solberg2018}. Therefore conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia in our study.\\

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. \textbf{TODO: Rewrite!} In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). To this end, we derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.
